\documentclass[../II2202-proposal.tex]{subfiles}

\begin{document}

\textbf{Informed Consent Form}

The following is an informed consent document for the purpose of freeing students/faculty of any liability while performing a research study with human participants. 
The study, performed by Stefano Formicola and Christoph Albert Johns, aims at illuminating the relationship between immersion and energy impact in deploying advanced augmented reality (AR) features, thus empowering developers and designers to decide the most appropriate feature sets for their applications considering both user experience and sustainability aspects.

The experiment will consist of two three-minute session of an Augmented Reality game, with and without a specific setting enabled, and data collection and measurement. Following, you will find a more precise description of collected data. \\

All information you provide will remain confidential and will not be associated with your name. If for any reason during this study you do not feel comfortable, you may leave the experiment at any time and your information will be discarded.
Your participation in this study will require approximately 30 minutes. When this study is complete you will be provided with the results of the experiment if you request them, and you will be free to ask any questions. \\

If you have any further questions concerning this study please feel free to contact us through email: Stefano Formicola at formico@kth.se or Christoph Albert Johns at cajohns@kth.se.
Your participation is solicited, yet strictly voluntary. All information will be kept confidential and your name will not be associated with any research findings and is only recorded to remove your data if so required. You can withdraw your consent at any time.

\begin{enumerate}
    \item Subject\\
    \textit{Name and Surname of the subject involved in the experiment. This will not be shared nor will the data be associated with your name.}
    \item Date\\
    \textit{DD/MM/YYYY}
    \item I agree to share: (checkbox)\\
    -- \textit{Device measurements (i.e. CPU usage, overall duration) during my usage }\\
    -- \textit{My questionnaire answers}
    \item Research study introduction (checkbox)\\
    -- \textit{I agree the experiment has been fully explained to me, as well as the data collected and my right to withdraw in any moment.}
    \item Research study consent (checkbox)\\
    -- \textit{I have read all the information of this form and am willing to participate in the research study.}
\end{enumerate}

\end{document}